The Timing and Likelihood of the Origin of Life from Earth's Impact History

### 2023-04-25 ms.tex
caveats in results and wordsmithing

\citet{Benner_2020} argued that the most likely time for the origin of life in an RNA-world scenario would have been $4.36 \pm 0.1$ Ga in the wake of a $\sim 2 \times 10^{22}$ kg ($\sim 2100$ km) asteroid impact which they called Moneta. They suggested that iron delivered by Moneta's core would have reacted with impact-vaporized ocean water to generate a reducing Hadean atmosphere conducive to the photochemical generation of essential prebiotic molecules like HCN. This transient period could have been a window of opportunity for the origin of RNA and, ultimately, life.

@@

Recently, \citep{Wogan_2023} use photochemical models of post-impact atmospheres to show that impacts significantly smaller than Moneta can efficiently produce prebiotic molecules. In their simulations, impactor iron equilibrates with vaporized ocean water to generate atmospheric H$_2$, and CH$_4$ and NH$_3$ form thermochemically as the reducing steam atmosphere cools. Once steam condense to an ocean, subsequent photochemistry of the Titan-like atmosphere produces prebiotic nitriles. Modeling shows that the production of both HCN and HCCCN only occurs when $\mathrm{CH_4} / \mathrm{CO_2} \gtrsim 0.1$ while the atmosphere is hazy. In an optimistic scenario, which includes nickel catalyzed methane production, modeling suggests that $\mathrm{CH_4} / \mathrm{CO_2} > 0.1$ (i.e. significant nitriles) occurs for impacts $> 4 \times 10^{20}$ kg. In a pessimistic modeling case, which assumes only a fraction of impactor iron reacts with the atmosphere \citep{Citron_2022}, a $> 5 \times 10^{21}$ kg impactor is required to produce substantial HCN and HCCCN. Regardless of uncertainty, \citet{Wogan_2023} show that an impactor much smaller than Moneta could spark an origin of life, and that the \citet{Benner_2020} estimated timing for biopoesis could be inaccurate because it is based on Moneta.

Here, we use the \citet{Wogan_2023} results, along with Monte-Carlo simulations of Earth's impact history to make an alternative estimate for when life most likely emerged during an RNA-first process. Our calculations account for the possibility of planet sterilization by impacts that vaporize the ocean \citep{Sleep_1989}. We assume that a life-starting impact is one that produces significant prebiotic molecules and is not subsequently followed by an ocean-vaporizing impact that destroys the biosphere. By considering the fraction of stochastic impact realizations that do not have an life-starting impact, we also estimate the probability of life beginning if Earth's clock was rerun.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.65\textwidth]{figures/example_impact_histories.pdf}
  \caption{Three simulated impact histories of the 5000 derived from our Monte-Carlo approach (Section \ref{sec:methods}). Each vertical line indicates an impact of a mass shown by the y-axis.}
  \label{fig:example_impact_histories}
\end{figure}

Figure \ref{fig:example_impact_histories} illustrates three of the 5000 impact histories computed with our Monte-Carlo approach (Section \ref{sec:methods}). In Figure \ref{fig:example_impact_histories}a, a Moneta-sized impact occurs at $\sim 4.43$ Ga delivering nearly all of Earth's mantle HSEs. The impact should also reduce the Hadean atmosphere creating conditions favorable for prebiotic chemistry. However, in this case, Moneta is not a life-starting impact because Moneta is followed by an ocean-vaporizing collision at $\sim 4.37$ Ga which we assume destroys any existing biosphere. Based on smoothed-particle hydrodynamics impact simulations, we nominally assume that 10\% of an impact's kinetic energy goes into vaporizing the ocean \citep[Appendix X,][]{Citron_2022}. Vaporizing the ocean requires $5 \times 10^{27}$ J, so a $> 5 \times 10^{28}$ J impact should deliver sufficient energy. In Figure \ref{fig:example_impact_histories}a, the last ocean-vaporizing impact is only $\sim 10^{20}$ kg (360 km) but has a large impact velocity of $32$ km/s giving it $5.12 \times 10^{28}$ J of kinetic energy.

@@

Figure \ref{fig:example_impact_histories}c shows a impact history where Moneta does not occur. Instead, Earth's mantle HSEs are delivered by multiple $10^{21}$ to $4 \times 10^{21}$ kg collisions. Whether this bombardment history would cause biopoesis depends on the required minimum impact mass to produce significant origin of life molecules. \citet{Wogan_2023} uses photochemical models of post-impact atmospheres to show that, optimistically, an impact $> 4 \times 10^{20}$ kg should give rise to an atmosphere that makes significant prebiotic molecules. Under this optimistic requirement, the last ocean-vaporizing impact in Figure \ref{fig:example_impact_histories}c at 4.38 Ga with mass $3 \times 10^{21}$ kg could successfully start life. However, with pessimistic modeling assumptions, \citet{Wogan_2023} finds that a $> 5 \times 10^{21}$ kg impact is instead required for biopoesis. In this alternative scenario, an origin of life never occurs in Figure \ref{fig:example_impact_histories}c because there is no impact larger than $5 \times 10^{21}$ kg.

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{figures/probabilities_of_impacts.pdf}
  \caption{The probability of an impact-driven origin of life on Earth. The blue solid line is the probability of an impact occurring at least once. The orange dashed line is the probability of an impact without a subsequent ocean-vaporizing collision, which can be interpreted as the chance of an impact that starts life. Probabilities are shown as a function of the minimum impact mass to produce significant prebiotic molecules. Two plausible minimum masses are the \citet{Wogan_2023} optimistic and pessimistic scenarios indicated with vertical dotted lines. The shaded region between $2 \times 10^{22}$ and $6 \times 10^{22}$ kg are Moneta-sized impacts because they deliver most all of Earth's mantle HSEs. There is a 92\% chance of a life-starting impact in the \citet{Wogan_2023} optimistic case, and a 41\% chance in the pessimistic case.}
  \label{fig:probabilities_of_impacts}
\end{figure}

As illustrated in Figure \ref{fig:example_impact_histories}, a life-starting impact does not occur in every simulated impact history. In our impact-driven model for the origin of life, biopoesis requires (1) an impact of sufficient mass to make prebiotic molecules and (2) the lack of a later ocean-vaporizing collision that destroys the biosphere without rekindling it. Figure \ref{fig:probabilities_of_impacts} shows the probability of conditions (1) and (2) occurring as a function of the minimum impact mass that produces prebiotic molecules. For example, consider the \citet{Wogan_2023} optimistic minimum mass of $4 \times 10^{20}$ kg indicated in Figure \ref{fig:probabilities_of_impacts}. All simulated impact histories have at least one impact bigger than $4 \times 10^{20}$ kg (blue line in Figure \ref{fig:probabilities_of_impacts}). However, the probability of the origin of life in this scenario is 92\%, because 8\% of the time the last $4 \times 10^{20}$ kg collision is followed by a smaller ocean-vaporizing impact that perhaps sterilizes the planet (orange dashed line in Figure \ref{fig:probabilities_of_impacts}). An origin of life is less probable if we consider the \citet{Wogan_2023} pessimistic minimum mass ($> 5 \times 10^{21}$ kg) to produce significant origin of life molecules. A $> 5 \times 10^{21}$ kg impact without later ocean vaporization occurs only 41\% of the time.

@@

% A caveat to the results above is that ocean-vaporization by an impact is likely a complicated function of impact mass, velocity, and incident angle, which we do not account for (Appendix \ref{sec:append_vap}). Through this section we have assumed that 10\% of an impactor's kinetic energy heats the planetary surface ($f_\mathrm{E,vap} = 0.1$), leading to a required $> 5 \times 10^{28}$ J collision to vaporize an ocean. Appendix Figure \ref{fig:energy_for_ocean_vap} uses simulations of impacts from \citet{Citron_2022} to show that $f_\mathrm{E,vap} = 0.1$ is a reasonable assumption, but that values between 2.5\% and 25\% are also possible and that $f_\mathrm{E,vap}$ likely also depends on impact angle.

@@

We use Monte-Carlo simulations of Earth's impact history to determine the most likely timing for the origin of life in an RNA-first processes. We make use of results of \citep{Wogan_2023}, who find that significant origin of life molecules, such as HCN, are produced in the Hadean atmosphere after a $> 4 \times 10^{20}$ kg impact in an optimistic case, and after a $> 5 \times 10^{21}$ kg impact in a pessimistic scenario. We consider both possibilities, and assume such impacts can cause life's origin so long as they are not followed by a smaller ocean-vaporizing collision that sterilizes the planet.

@@

% Caveats
% While the modeling effort described above attempts to use the most current understanding of Earth's impact history and its effects,

@@

>>>>

\citet{Benner_2020} argued that the most likely time for the origin of life in an RNA-world scenario would have been $4.36 \pm 0.1$ Ga in the wake of a $\sim 2 \times 10^{22}$ kg ($\sim 2100$ km) asteroid impact which they called Moneta. They suggested that iron delivered by Moneta's core would have reacted with impact-vaporized ocean water to generate a reducing Hadean atmosphere conducive to the photochemical generation of essential prebiotic nitriles, like HCN and HCCCN. Nitriles are required in prebiotic schemes that synthesize ribonucleotide precursors to RNA \citep{Patel_2015,Powner_2009,Becker_2019}. A transient period of rampant nitrile production after a Moneta impact could have been a window of opportunity for the origin of RNA and, ultimately, life.

@@

Recently, \citep{Wogan_2023} use photochemical models of post-impact atmospheres to show that impacts significantly smaller than Moneta can efficiently produce prebiotic molecules. In their simulations, impactor iron equilibrates with vaporized ocean water to generate atmospheric H$_2$, and CH$_4$ and NH$_3$ form thermochemically as the reducing steam atmosphere cools. Once steam condense to an ocean, subsequent photochemistry of the Titan-like atmosphere produces prebiotic nitriles. Modeling shows that the production of both HCN and HCCCN only occurs when $\mathrm{CH_4} / \mathrm{CO_2} \gtrsim 0.1$ while the atmosphere is hazy. In an optimistic scenario, which includes nickel-catalyzed methane production, modeling suggests that $\mathrm{CH_4} / \mathrm{CO_2} > 0.1$ (i.e. significant nitriles) occurs for impacts $> 4 \times 10^{20}$ kg. In a pessimistic modeling case, which assumes only a fraction of impactor iron reacts with the atmosphere \citep{Citron_2022}, a $> 5 \times 10^{21}$ kg impactor is required to produce substantial HCN and HCCCN. Regardless of uncertainty, \citet{Wogan_2023} show that an impactor much smaller than Moneta could deliver the nitriles needed for an RNA origin of life, and that the \citet{Benner_2020} estimated timing for biopoesis could be inaccurate because it is based on Moneta.

Here, we use the \citet{Wogan_2023} results, along with Monte-Carlo simulations of Earth's impact history to make an alternative estimate for when life most likely emerged during an RNA-first process. Our calculations account for the possibility of planet sterilization by impacts that vaporize the ocean \citep{Sleep_1989}. We assume that a life-starting impact is one that produces significant prebiotic nitriles and is not subsequently followed by an ocean-vaporizing impact that destroys the biosphere. By considering the fraction of stochastic impact realizations that do not have an life-starting impact, we also estimate the probability of life beginning if Earth's clock was rerun.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.65\textwidth]{figures/example_impact_histories.pdf}
  \caption{Three simulated impact histories of the 5000 derived from our Monte-Carlo approach (Section \ref{sec:methods}). Each vertical line indicates an impact of a mass shown by the y-axis. The red lines indicate the last impact to vaporize the ocean assuming 10\% of an impacts kinetic energy heats the ocean.}
  \label{fig:example_impact_histories}
\end{figure}

Figure \ref{fig:example_impact_histories} illustrates three of the 5000 impact histories computed with our Monte-Carlo approach (Section \ref{sec:methods}). In Figure \ref{fig:example_impact_histories}a, a Moneta-sized impact occurs at $\sim 4.43$ Ga delivering nearly all of Earth's mantle HSEs. The impact should also reduce the Hadean atmosphere creating conditions favorable for prebiotic chemistry. However, in this case, Moneta is not a life-starting impact because Moneta is followed by an ocean-vaporizing collision at $\sim 4.37$ Ga which we assume destroys any existing biosphere. Based on smoothed-particle hydrodynamics impact simulations, we nominally assume that 10\% of an impact's kinetic energy goes into vaporizing the ocean \citep[Appendix \ref{sec:append_vap},][]{Citron_2022}. Vaporizing the ocean requires $5 \times 10^{27}$ J, so a $> 5 \times 10^{28}$ J impact should deliver sufficient energy under our assumptions. Appendix \ref{sec:append_vap} considers alternative criteria for ocean-vaporization which we discuss at the end of the Results section. In Figure \ref{fig:example_impact_histories}a, the last ocean-vaporizing impact is only $\sim 10^{20}$ kg (360 km) but has a large impact velocity of $32$ km/s giving it $5.12 \times 10^{28}$ J of kinetic energy.

@@

Figure \ref{fig:example_impact_histories}c shows a impact history where Moneta does not occur. Instead, Earth's mantle HSEs are delivered by multiple $10^{21}$ to $4 \times 10^{21}$ kg collisions. Whether this bombardment history would cause biopoesis depends on the required minimum impact mass to produce significant origin of life nitriles (e.g. HCN and HCCCN). \citet{Wogan_2023} uses photochemical models of post-impact atmospheres to show that, optimistically, an impact $> 4 \times 10^{20}$ kg should give rise to an atmosphere that makes significant HCN and HCCCN. Under this optimistic requirement, the last ocean-vaporizing impact in Figure \ref{fig:example_impact_histories}c at 4.38 Ga with mass $3 \times 10^{21}$ kg could successfully start life. However, with pessimistic modeling assumptions, \citet{Wogan_2023} finds that a $> 5 \times 10^{21}$ kg impact is instead required for substantial nitrile delivery to the surface. In this alternative scenario, an origin of life never occurs in Figure \ref{fig:example_impact_histories}c because there is no impact larger than $5 \times 10^{21}$ kg.

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{figures/probabilities_of_impacts.pdf}
  \caption{The probability of an impact-driven origin of life on Earth. The blue solid line is the probability of an impact occurring at least once. The orange dashed line is the probability of an impact without a subsequent ocean-vaporizing collision, which can be interpreted as the chance of an impact that starts life. Probabilities are shown as a function of the minimum impact mass to produce significant prebiotic molecules (e.g. HCN). Two plausible minimum masses are the \citet{Wogan_2023} optimistic and pessimistic scenarios indicated with vertical dotted lines. The shaded region between $2 \times 10^{22}$ and $6 \times 10^{22}$ kg are Moneta-sized impacts because they deliver most all of Earth's mantle HSEs. There is a 92\% chance of a life-starting impact in the \citet{Wogan_2023} optimistic case, and a 41\% chance in the pessimistic case.}
  \label{fig:probabilities_of_impacts}
\end{figure}

As illustrated in Figure \ref{fig:example_impact_histories}, a life-starting impact does not occur in every simulated impact history. In our impact-driven model for the origin of life, biopoesis requires (1) an impact of sufficient mass to make prebiotic nitriles and (2) the lack of a later ocean-vaporizing collision that destroys the biosphere without rekindling it. Figure \ref{fig:probabilities_of_impacts} shows the probability of conditions (1) and (2) occurring as a function of the minimum impact mass that produces prebiotic molecules. For example, consider the \citet{Wogan_2023} optimistic minimum mass of $4 \times 10^{20}$ kg indicated in Figure \ref{fig:probabilities_of_impacts}. All simulated impact histories have at least one impact bigger than $4 \times 10^{20}$ kg (blue line in Figure \ref{fig:probabilities_of_impacts}). However, the probability of the origin of life in this scenario is 92\%, because 8\% of the time the last $4 \times 10^{20}$ kg collision is followed by a smaller ocean-vaporizing impact that perhaps sterilizes the planet (orange dashed line in Figure \ref{fig:probabilities_of_impacts}). An origin of life is less probable if we consider the \citet{Wogan_2023} pessimistic minimum mass ($> 5 \times 10^{21}$ kg) to produce significant origin of life nitriles. A $> 5 \times 10^{21}$ kg impact without later ocean vaporization occurs 41\% of the time. A life-starting impact was therefore not an anomalous event on the early Earth, and was strongly favored (92\% probability) under optimistic assumptions.

@@

In this model for biopoesis, life would start approximately 10s of millions of years (i.e., 0.01s of billions of years) after a life-starting impact because this is the maximum duration of the H$_2$- and CH$_4$- rich atmosphere that makes nitriles like HCN \citep{Wogan_2023}. We ignore the $< 10$s million year span between impact and life's origin because it is small compared to our $\sim 500$ million-year uncertainty for the timing of a life-starting impact.

A caveat to the results above is that ocean-vaporization by an impact is likely a complicated function of impact mass, velocity, and incident angle, which we do not account for (Appendix \ref{sec:append_vap}). Through this section we have assumed that 10\% of an impactor's kinetic energy heats the planetary surface ($f_\mathrm{E,vap} = 0.1$), leading to a required $> 5 \times 10^{28}$ J collision to vaporize an ocean. Appendix Figure \ref{fig:energy_for_ocean_vap} uses simulations of impacts from \citet{Citron_2022} to show that $f_\mathrm{E,vap} = 0.1$ is a reasonable assumption, but that values between 2.5\% and 25\% are also possible and that $f_\mathrm{E,vap}$ likely also depends on impact angle.

To determine the sensitivity of our results to $f_\mathrm{E,vap}$, we recomputed the likelihood and timing of an impact that starts life in Appendix \ref{sec:append_vap} using $f_\mathrm{E,vap} = 0.025$ and $f_\mathrm{E,vap} = 0.25$. We find that the probability of a life-starting impact is somewhat sensitive to $f_\mathrm{E,vap}$, while the timing does not depend strongly on $f_\mathrm{E,vap}$. Overall, this uncertainty does not change our qualitative conclusion that an origin of life impact on the early Earth would not be a fluke. Even in our worst-case scenario (i.e., \citet{Wogan_2023} pessimistic case with $f_\mathrm{E,vap} = 0.25$) a life-starting impact occurs in 20\% of impact realizations (Appendix Figure \ref{fig:probabilities_of_impacts_sens}). Additionally, uncertainty in $f_\mathrm{E,vap}$ does not change our general conclusion that the most likely timing for the origin of life was $\sim 4.35$ Ga, with 95\% uncertainty spanning the Hadean eon (Appendix Figure \ref{fig:timing_sensitivity}). However, a more complete understanding of $f_\mathrm{E,vap}$ as a function of impact angle and other similar parameters may change this result.

% The timing of the last life-starting impact was not very sensitive to $f_\mathrm{E,vap}$ (Appendix Figure \ref{fig:timing_sensitivity}). Smaller $f_\mathrm{E,vap}$ favored a life-starting impactor earlier in the Hadean because small $f_\mathrm{E,vap}$ reduces the chance of late ocean-vaporization. Regardless, the median time of impact is always between 4.29 and 4.41 Ga which 95\% uncertainties spanning the Hadean eon.

% The probability of an impact that produces prebiotic molecules without later ocean vaporization (i.e., a life-starting impact) is somewhat sensitive to $f_\mathrm{E,vap}$ (Appendix Figure \ref{fig:probabilities_of_impacts_sens}). Smaller $f_\mathrm{E,vap}$ makes ocean-vaporization more rare which increases the chance of a life-starting impact.

@@

We use Monte-Carlo simulations of Earth's impact history to determine the most likely timing for the origin of life in an RNA-first processes. We make use of results of \citet{Wogan_2023}, who find that significant origin of life molecules, such as HCN, are produced in the Hadean atmosphere after a $> 4 \times 10^{20}$ kg impact in an optimistic case, and after a $> 5 \times 10^{21}$ kg impact in a pessimistic scenario. We consider both possibilities, and assume such impacts can cause life's origin so long as they are not followed by a smaller ocean-vaporizing collision that sterilizes the planet.

@@

@@

% Ocean vaporization caveat Caveat.
